Analyzing the sensitivity and specificity of ASVs for discriminating between genomes

### 2020-09-16 exploratory/2020-09-09-genome-sens-spec.Rmd
Combine the count_tibble files and update genome-level exploratory analysis, closes #23

---
title: "Analyzing the sensitivity and specificity of ASVs for discriminating between genomes"
author: "Pat Schloss"
date: "9/9/2020"
output:
  github_document:
    html_preview: false
---

@@

```{r}
fl <- read_tsv(here("data/v19/rrnDB.count_tibble"))
```

@@

```{r n_rrn}
fl %>%
	group_by(genome) %>%
	summarize(n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn)) + geom_histogram(binwidth=1)

fl %>%
	group_by(genome) %>%
	summarize(n_rrn = sum(count), .groups="drop") %>%
	count(n_rrn) %>%
	mutate(fraction = n / sum(n))
```

@@

```{r}
fl %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(n_rrn) %>%
	summarize(med_n_asv = median(n_asv),
						lq_n_asv = quantile(n_asv, prob=0.25),
						uq_n_asv = quantile(n_asv, prob=0.75))

fl %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_asv)) + geom_smooth(method="lm")
```

Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome.

@@

```{r}
fl %>%
	group_by(asv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n))
```

We see that will full length sequences, that 82% of the ASVs were unique to a genome.

### Does the sensitivity and specificity change if we look at a shorter region of the gene?

We know that the V4 region is less diverse than the full length sequence. So does the number of ASVs per genome differ than for full length sequences? Are ASVs as specific when using the V4 region compared to full length sequences?

```{r}
v_four <- read_tsv(here("data/v4/rrnDB.count_tibble"))

v_four %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(n_rrn) %>%
	summarize(mean_n_asv = mean(n_asv),
						lq_n_asv = quantile(n_asv, prob=0.25),
						uq_n_asv = quantile(n_asv, prob=0.75))

v_four %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_asv)) + geom_smooth(method="lm")
```

The number of ASVs per copy of *rrn* operon is lower than for full length sequences. We find 1.5 ASV per 10 copies of the *rrn* operon.

Next, let's look at the specificity of an ASV for a genome

```{r}
v_four %>%
	group_by(asv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n))
```

We find that about 76% of ASVs using the V4 region were specific to a genome.

>>>>

---
title: "Analyzing the sensitivity and specificity of ASVs for discriminating between genomes"
author: "Pat Schloss"
date: "9/9/2020"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

@@

```{r}
count_tibble <- read_tsv(here("data/processed/rrnDB.count_tibble"),
												 col_types = "cccd")
```

@@

```{r n_rrn}
count_tibble %>%
	filter(region == "v19") %>%
	group_by(genome) %>%
	summarize(n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn)) + geom_histogram(binwidth=1)

count_tibble %>%
	filter(region == "v19") %>%
	group_by(genome) %>%
	summarize(n_rrn = sum(count)) %>%
	count(n_rrn) %>%
	mutate(fraction = n / sum(n))
```

@@

```{r}
count_tibble %>% 
	group_by(region, genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(region, n_rrn) %>%
	summarize(med_n_asv = median(n_asv),
						mean_n_asv = mean(n_asv),
						lq_n_asv = quantile(n_asv, prob=0.25),
						uq_n_asv = quantile(n_asv, prob=0.75)) %>%
	filter(n_rrn == 7)

count_tibble %>% 
	group_by(region, genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_asv, color=region)) + geom_smooth(method="lm")
```

Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome. The sub regions of the 16S rRNA region have few ASVs per *rrn* operon.

@@

```{r}
count_tibble %>%
	group_by(region, asv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n)) %>%
	filter(n_genomes == 1)
```

We see that will full length sequences, that 82% of the ASVs were unique to a genome. For the subregions, about 76% of the ASVs were unique to a genome.
